Embedding Intuitionistic into Classical Logic

### 2023-03-09 LPAR/LPAR-EIICL.tex
comments appendix

\begin{proof}
	1. If $C^I(m)$ is true then we are done due to persistency. Otherwise $C^I(m)$ is false.
	Because $(A^I(f_{[\psi]}^I(m))\to B^I(f_{[\psi]}^I(m)))\to C^I(u)$ holds, then $A^I(f_{[\psi]}^I(u))\to B^I(f_{[\psi]}^I(u))$ must be false.

@@

\begin{definition}
For every propositional variable $A$ occurring in $\varphi$ and non-negative integer $n$ (representing the distance of a node from the root on a path) consider a new propositional variable $A^n$ and for every formula $\psi\in\mathcal X$ consider a new propositional variable $X_\psi^n$.
Let $\vec A^n$ range over all propositional variables $A^n$ and $\vec X^n$ range over all $X_\psi^n$. 
Define
	\begin{itemize}
		\item {\normalfont Valid}$(n)$ which encodes that $\vec X^n_\psi$ represents a valid next element of a sequence without repetition, i.e. if we interpret the $\vec X^i$ as bit-vectors, then $\vec X^n$ has exactly one bit set to $1$, indicating the formula $\psi$ at the $n$-th position of the sequence, and it is at a position that is $0$ in all $\vec X^i$ with $i < n$ (i.e., it did not occur in the sequence previously).
		\item {\normalfont Persistent}$(n)$ which encodes that the persistency condition holds.
		\item {\normalfont Sat}$_{\mathcal R}(n)$ encoding that the formulas in $\mathcal R$ hold at the $n$-th world of the path.
		\item {\normalfont Sat}$_{\mathcal X}(n)$ encoding that the formula in $\mathcal X$ that is represented by $\vec X^{n-1}$ holds.
	\end{itemize}
	Define$$\varphi^Q_{i} = \exists \vec A^i\forall \vec X^i\left(\text{\normalfont Persistent}(i)\wedge \text{\normalfont Sat}_{\mathcal R}(i)\wedge \text{\normalfont Sat}_{\mathcal X}(i)\wedge\left(\text{\normalfont Valid}(i)\to \varphi^Q_{i+1}\right)\right)$$
	for $0 < i < |\mathcal{X}|$, as well as the special cases
	$$\varphi^Q_{|\mathcal{X}|} = \exists \vec A^{|\mathcal{X}|}\left(\text{\normalfont Persistent}(|\mathcal{X}|)\wedge \text{\normalfont Sat}_{\mathcal R}(|\mathcal{X}|)\wedge \text{\normalfont Sat}_{\mathcal X}(|\mathcal{X}|)\right)$$
	for leaves, and $$\varphi^Q = \varphi_0^Q = \exists \vec A^0\forall \vec X^0\left(\neg P^0\wedge \text{\normalfont Sat}_{\mathcal R}(0)\wedge(\text{\normalfont Valid}(0)\to \varphi_1^Q)\right)$$for the root.
	Example encodings of the above formulas are:
	\begin{align*}
		\text{\normalfont Valid}(n) = &\left(\bigvee\left\{X^n_\psi\:|\:\psi\in\mathcal X\right\}\right)\wedge\left(\bigwedge \left\{\neg(X^n_{\psi_1}\wedge X^n_{\psi_2})\:|\:\psi_1\neq\psi_2\in\mathcal X\right\}\right)\wedge\\&\left(\bigwedge\left\{\bigwedge\{X^i_\psi\to\neg X^n_\psi\:|\:i < n\}\:|\:\psi\in\mathcal X\right\}\right)\\
		\text{\normalfont Persistent}(n) = & \bigwedge\left\{A^{n-1}\to A^n\:|\: A \text{ prop. variable with }A^{n-1}\in\vec A^{n-1}, A^n \in\vec A^{n}\right\}\\
		\text{\normalfont Sat}_{\mathcal R}(n) = &\bigwedge\left\{\bigwedge A^n_i\to\bigvee B^n_j\:|\:\bigwedge A_i\to\bigvee B_j\mathcal R\right\}\\
		\text{\normalfont Sat}_{\mathcal X}(n) = & \bigwedge\left\{X^{n-1}_\psi\to (A^n\to B^n)\to C^{n-1}\:|\:\psi = (A\to B)\to C\in\mathcal X\right\}
	\end{align*}
\end{definition}

@@

\begin{lemma}\label{lemma:QBF}
	$\varphi$ is not intuitionistically valid iff $\varphi^Q$ is a satisfiable QBF.
\end{lemma}
\begin{proof}
Suppose $\varphi$ is not intuitionistically valid.
Then, $(\bigwedge\mathcal R\wedge\bigwedge\mathcal X)\to P$ has an intuitionistic counter-model. By the previous section there exists a classical counter-model $\mathcal M$ for $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$.
We now define a QBF model iteratively.
For each atom $A$ interpret $A^0$ such as $\mathcal M$ interprets $A^\epsilon$. Suppose we are given interpretations of all atoms $A^i$ for $i < n$ and a sequence with no repetitions $\psi_1\dots\psi_{n-1}$ over $\mathcal X$ such that $\psi_i$ is exactly the $\psi\in\mathcal X$ for which $X_{\psi}^i$ is true and $A^i$ is interpreted as $A^{\psi_1\dots\psi_i}$  in $\mathcal M$.
Let the $X^{n}_\psi$ be arbitrarily interpreted (since they are $\forall$-quantified). If not exactly one of the $X^{n}_\psi$ is interpreted as true, then valid$(n)$ fails and the remaining propositions can be chosen arbitrarily. If $X^i_{\psi_n} = 1$ for some $i < n$, then valid$(n)$ also fails and the remaining propositions can be chosen arbitrarily.
So we may assume that $\psi_1\dots\psi_n$ is a sequence with no repetitions.
Interpret the atoms $A^n$ as $\mathcal M$ interprets $A^{\psi_1\dots\psi_n}$.
Continue this construction until $n  = |\mathcal X|$. Then from $\mathcal M$ being a counter-example to $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$ it directly follows that this interpretation satisfies $\varphi^Q$.

@@

	\emph{Claim.} For any formula $\psi$ in which the free variable $b$ doesn't occur, world $w_k\in W_0$ and variable assignments $v_c, v_i$ such that $v_c(a) = [v_k(a), w_k]$ for all free variables $a$ occurring in $\psi$ it holds that \[M, I, v_c[w_k/b]\models \psi^b\text{ if and only if $w_k, v_i\models \psi$.}\]

@@

	For the induction step consider e.g. the case where $\psi$ is of the form $\forall x \psi'$.
	Suppose first that $w_k, v_i\not\models \forall x\psi'$, i.e. there exists $w_l\geq w_k$ and $m_l\in w_l$ such that $w_l, v_i[m_l/a]\not\models \psi'[a/x]$. Then by the induction hypothesis $M, I, v_c[[v_i(a), w_l],w_l/c]\not\models \psi'[a/x]^c$. Furthermore by definition $M, I, v_c[w_l/c, w_k/b]\models b\preceq c$ and $M, I, v_c[v_c[[v_i(a), w_l],w_l/c]\models E(a, c)$. Hence $M, I, v_c[w_k/b]\not\models \psi^b$. Suppose on the other hand that $w_k, v_i\not\models\forall x\psi'$, i.e. for all $w_l\geq w_k$ and $m_l\in w_l$ it holds that $w_l, v_i[m_l/a]\models \psi'[a/x]$. Then by the induction hypothesis $M, I, v_c[[m_l, w_l]/A,w_l/c]\models \psi'[a/x]^c$ for all $w_l\geq w_k$ and $m_l\in w_l$. It can be shown that for every other assignment of $c$ and $a$ either $b\preceq c$ or $E(a, c)$ is violated. Hence $M, I, v_c[w_k/b]\models \psi^b$. For other forms of $\psi$ one proceeds in similar fashion. This completes the proof of the claim.

	Recall that $w_0, v\not\models\varphi$. From the claim it follows that \[M, I, v'[[v(a_0), w_0]/a_0\dots[v(a_n), w_0]/a_0, w_0/b]\not\models \varphi^b\]where $a_0\dots a_n$ are the free variables occurring in $\varphi$ and $v'$ is any variable assignment. Furthermore a straightforward check of definitions yields \[M, I, v'[[v(a_0), w_0]/a_0\dots[v(a_n), w_0]/a_0, w_0/b]\models K(\varphi)\wedge \text{World}(b)\wedge\vec E(\vec a, b).\]Therefore $M, I\not\models \varphi^C$ and we are done.

\end{proof}

>>>>

\begin{proof}
	1. If $C^I(m)$ is true then we are done due to persistency. Otherwise $C^I(m)$ is false.
	Because $(A^I(f_{[\psi]}^I(m))\to B^I(f_{[\psi]}^I(m)))\to C^I(m)$ holds, then $A^I(f_{[\psi]}^I(m))\to B^I(f_{[\psi]}^I(m))$ must be false.

@@

\begin{definition}
For every propositional variable $A$ occurring in $\varphi$ and non-negative integer $n$ (representing the $n$-th position of a node on a path) consider a new propositional variable $A^n$ and for every formula $\psi\in\mathcal X$ consider a new propositional variable $X_\psi^n$.
Let $\vec A^n$ range over all propositional variables $A^n$ and $\vec X^n$ range over all $X_\psi^n$.
Define
	\begin{itemize}
		\item {\normalfont Valid}$(n)$ which encodes that $\vec X^n_\psi$ represents a valid next element of a sequence without repetition, i.e. if we interpret the $\vec X^i$ as bit-vectors, then $\vec X^n$ has exactly one bit set to $1$, indicating the formula $\psi$ at the $n$-th position of the sequence, and it is at a position that is $0$ in all $\vec X^i$ with $i < n$ (i.e., it did not occur in the sequence previously).
		\item {\normalfont Persistent}$(n)$ which encodes that the persistency condition holds.
		\item {\normalfont Sat}$_{\mathcal R}(n)$ encoding that the formulas in $\mathcal R$ hold at the $n$-th world of the path.
		\item {\normalfont Sat}$_{\mathcal X}(n)$ encoding that the formula in $\mathcal X$ that is represented by $\vec X^{n-1}$ holds.
	\end{itemize}
	Define$$\varphi^Q_{i} = \exists \vec A^i\forall \vec X^i\left(\text{\normalfont Persistent}(i)\wedge \text{\normalfont Sat}_{\mathcal R}(i)\wedge \text{\normalfont Sat}_{\mathcal X}(i)\wedge\left(\text{\normalfont Valid}(i)\to \varphi^Q_{i+1}\right)\right)$$
	for $0 < i < |\mathcal{X}|$, as well as the special cases
	$$\varphi^Q_{|\mathcal{X}|} = \exists \vec A^{|\mathcal{X}|}\left(\text{\normalfont Persistent}(|\mathcal{X}|)\wedge \text{\normalfont Sat}_{\mathcal R}(|\mathcal{X}|)\wedge \text{\normalfont Sat}_{\mathcal X}(|\mathcal{X}|)\right)$$
	for leaves, and $$\varphi^Q = \varphi_0^Q = \exists \vec A^0\forall \vec X^0\left(\neg P^0\wedge \text{\normalfont Sat}_{\mathcal R}(0)\wedge(\text{\normalfont Valid}(0)\to \varphi_1^Q)\right)$$for the root.
	Example encodings of the above formulas are:
	\begin{align*}
		\text{\normalfont Valid}(n) = &\left(\bigvee\left\{X^n_\psi\:|\:\psi\in\mathcal X\right\}\right)\wedge\left(\bigwedge \left\{\neg(X^n_{\psi_1}\wedge X^n_{\psi_2})\:|\:\psi_1\neq\psi_2\in\mathcal X\right\}\right)\wedge\\&\left(\bigwedge\left\{\bigwedge\{X^i_\psi\to\neg X^n_\psi\:|\:i < n\}\:|\:\psi\in\mathcal X\right\}\right)\\
		\text{\normalfont Persistent}(n) = & \bigwedge\left\{A^{n-1}\to A^n\:|\: A \text{ prop. variable with }A^{n-1}\in\vec A^{n-1}, A^n \in\vec A^{n}\right\}\\
		\text{\normalfont Sat}_{\mathcal R}(n) = &\bigwedge\left\{\bigwedge A^n_i\to\bigvee B^n_j\:|\:\bigwedge A_i\to\bigvee B_j \in \mathcal R\right\}\\
		\text{\normalfont Sat}_{\mathcal X}(n) = & \bigwedge\left\{X^{n-1}_\psi\to (A^n\to B^n)\to C^{n-1}\:|\:\psi = (A\to B)\to C\in\mathcal X\right\}
	\end{align*}
\end{definition}

@@

\begin{lemma}\label{lemma:QBF}
	$\varphi$ is not intuitionistically valid iff $\varphi^Q$ is a satisfiable QBF.
\end{lemma}
\begin{proof}
Suppose $\varphi$ is not intuitionistically valid.
Then, $(\bigwedge\mathcal R\wedge\bigwedge\mathcal X)\to P$ has an intuitionistic counter-model. By the previous section\todo{unklar was mit "previous section" gemeint ist, und was für ein model du genau meinst.} there exists a classical counter-model $\mathcal M$ for $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$.
We now define a QBF model iteratively.
For each atom $A$ interpret $A^0$ such as $\mathcal M$ interprets $A^\epsilon$. Suppose we are given interpretations of all atoms $A^i$ for $i < n$ and a sequence with no repetitions $\psi_1\dots\psi_{n-1}$ over $\mathcal X$ such that $\psi_i$ is exactly the $\psi\in\mathcal X$ for which $X_{\psi}^i$ is true and $A^i$ is interpreted as $A^{\psi_1\dots\psi_i}$  in $\mathcal M$.
Let the $X^{n}_\psi$ be arbitrarily interpreted (since they are $\forall$-quantified). If not exactly one of the $X^{n}_\psi$ is interpreted as true, then valid$(n)$ fails and the remaining propositions can be chosen arbitrarily. If $X^i_{\psi_n} = 1$ for some $i < n$, then valid$(n)$ also fails and the remaining propositions can be chosen arbitrarily.
So we may assume that $\psi_1\dots\psi_n$ is a sequence with no repetitions.
Interpret the atoms $A^n$ as $\mathcal M$ interprets $A^{\psi_1\dots\psi_n}$.
Continue this construction until $n  = |\mathcal X|$. Then from $\mathcal M$ being a counter-example to $(\bigwedge\mathcal R^\#\wedge\bigwedge\mathcal X^\#)\to P^\epsilon$ it directly follows that this interpretation satisfies $\varphi^Q$.

@@

	\emph{Claim.} For any formula $\psi$ in which the free variable $b$ doesn't occur, world $w_k\in W_0$ and variable assignments $v_c, v_i$ such that $v_c(a) = [v_k(a), w_k]$ \todo{das sollte $v_i$ statt $v_k$ sein, oder} for all free variables $a$ occurring in $\psi$ it holds that
\todo{Hier und weiter und sind immer M,I genannt, aber eigentlich haben wir Notation für $\mathcal M,v \models$ definiert}
\[M, I, v_c[w_k/b]\models \psi^b\text{ if and only if $w_k, v_i\models \psi$.}\]

@@

	For the induction step consider e.g. the case where $\psi$ is of the form $\forall x \psi'$.
	Suppose first that $w_k, v_i\not\models \forall x\psi'$, i.e. there exists $w_l\geq w_k$ and $m_l\in w_l$ such that $w_l, v_i[m_l/a]\not\models \psi'[a/x]$. Then by the induction hypothesis $M, I, v_c[[v_i(a), w_l],w_l/c]\not\models \psi'[a/x]^c$\todo{hier fehlt die Ersetzung für $[v_i(a), w_l]$. Fehlt außerdem nicht irgendein Bezug zu $m_l$?}. Furthermore by definition $M, I, v_c[w_l/c, w_k/b]\models b\preceq c$ and $M, I, v_c[v_c[[v_i(a), w_l],w_l/c]\models E(a, c)$\todo{hier fehlt die Ersetzung für $[v_i(a), w_l]$. Außerdem ist hier wirklich $v_c[v_c$ gemeint?}. Hence $M, I, v_c[w_k/b]\not\models \psi^b$. Suppose on the other hand that $w_k, v_i\not\models\forall x\psi'$, i.e. for all $w_l\geq w_k$ and $m_l\in w_l$ it holds that $w_l, v_i[m_l/a]\models \psi'[a/x]$. Then by the induction hypothesis $M, I, v_c[[m_l, w_l]/A,w_l/c]\models \psi'[a/x]^c$\todo{Hier wohl a statt A} for all $w_l\geq w_k$ and $m_l\in w_l$. It can be shown that for every other assignment of $c$ and $a$ either $b\preceq c$ or $E(a, c)$ is violated. Hence $M, I, v_c[w_k/b]\models \psi^b$. For other forms of $\psi$ one proceeds in similar fashion. This completes the proof of the claim.

	Recall that $w_0, v\not\models\varphi$. From the claim it follows that \[M, I, v'[[v(a_0), w_0]/a_0\dots[v(a_n), w_0]/a_0, w_0/b]\not\models \varphi^b\]where $a_0\dots a_n$ are the free variables occurring in $\varphi$ and $v'$ is any variable assignment. Furthermore a straightforward check of definitions yields \[M, I, v'[[v(a_0), w_0]/a_0\dots[v(a_n), w_0]/a_0, w_0/b]\models K(\varphi)\wedge \text{World}(b)\wedge\vec E(\vec a, b).\]Therefore $M, I\not\models \varphi^C$ and we are done.
\todo{Da ich hier nicht durchgekommen bin, kannst du den ganzen Beweis nochmal checken :p ?}
\end{proof}
